Emergence of the Southeast Asian islands as a driver for Neogene cooling

### 2020-08-03 Manuscript/Manuscript.tex
revision checkpoint

The results of our ``paleo-SEAI'' scenarios highlight the importance of the combination of topography, runoff, and lithology in setting Earth's climate state. To independently explore the effect of the modern-day surface exposure of low-relief basaltic lavas on steady-state \pCOtwo \cite{Kent2013a}, we replace mafic volcanics associated with the Deccan Traps, Ethiopian Traps, and Columbia River Basalts with the Ca+Mg concentration of bulk continental crust in GEOCLIM (Fig. \ref{fig:scenario_pCO2}). The resulting \pCOtwo is $<$500~ppm, indicating that, while the presence of mafic rocks in these igneous provinces modulates steady-state \pCOtwo as has been suggested to be important for Paleogene cooling \cite{Kent2013a}, its contribution is less than that of the SEAI, which have high-relief and a wet tropical climate in conjunction with high Ca+Mg lithologies.

@@

Marine \SrSr has overall been increasing since ca. 35~Ma \cite{McArthur2012a}. The traditional explanation for this trend is that it reflects increased weathering of radiogenic (i.e. high \SrSr) silicate rocks \cite{Raymo1988a}. Associated with this explanation is the proposal that increasing weathering of radiogenic silicate rocks in the Himalayas was the primary driver of Neogene cooling \cite{Raymo1992a}. It could be argued that increasing marine \SrSr is inconsistent with the hypothesis that increasing weathering of juvenile (i.e. low \SrSr) silicate rocks in the SEAI was an important driver of Neogene cooling. However, the globally averaged ratio of silicate weathering fluxes from radiogenic cratonic rocks versus juvenile arc lithologies can be at least partially decoupled from marine \SrSr via the regional weathering of isotopically unique lithologies. For example, unusually radiogenic carbonates are abundant in Himalayan strata, and it is estimated that $\sim$75\% of Sr coming from the Himalayas can be attributed to carbonate rather than silicate weathering \cite{Jacobson2002a, Quade2003a, Oliver2003a}. As such, there are challenges in interpreting the marine \SrSr record as a direct proxy for silicate weathering fluxes. Nevertheless, steadily increasing marine \SrSr is interrupted ca. 16~Ma when the slope of the \SrSr curve decreases \cite{McArthur2012a}. This decrease in slope has been attributed to coincident exhumation of relatively low \SrSr Outer Lesser Himalaya carbonates \cite{Myrow2015a, Colleps2018a}, but could also be at least partially driven by the emergence of low \SrSr lithologies in the SEAI during arc-continent collision.

@@

The long-term carbon-cycle mass balance can be perturbed via mechanisms that are disconnected from changes in volcanic degassing and silicate weathering rates. For example, sulphide oxidation coupled to carbonate dissolution could act as a source of \COtwo on million year time-scales \cite{Torres2014a}. Similarly, the weathering of sedimentary organic matter could serve as a source of \COtwo \cite{Hilton2014a}. On the other hand, enhanced burial of organic matter could be an important sink of \COtwo, as has been suggested in the Bengal Fan \cite{Galy2007a} and Taiwan \cite{Kao2014a}. The fluxes of \COtwo represented by these processes are not accounted for in our model framework, and could have been affected by emergence of the SEAI and/or Himalayan orogeny. \pCOtwo changes that result from these processes would be superimposed on \pCOtwo changes associated with evolving silicate weathering rates. However, our coupled weathering-climate model indicates that the \pCOtwo change associated with evolving silicate weathering rates driven by emergence of the SEAI is sufficient to explain the majority of Neogene cooling (Fig. \ref{fig:scenario_pCO2}). Without this emergence, \pCOtwo would have remained above the level necessary for the growth of Northern Hemisphere ice sheets.

>>>>

The results of our ``paleo-SEAI'' scenarios highlight the importance of the combination of topography, runoff, and lithology in setting Earth's climate state. To independently explore the effect of the modern-day surface exposure of lower-relief basaltic lavas on steady-state \pCOtwo \cite{Kent2013a}, we replace mafic volcanics associated with the Deccan Traps, Ethiopian Traps, and Columbia River Basalts with the Ca+Mg concentration of bulk continental crust in GEOCLIM (Fig. \ref{fig:scenario_pCO2}). The resulting \pCOtwo is $\sim$300--500~ppm, indicating that the presence of mafic rocks in these igneous provinces affects steady-state \pCOtwo as has been suggested to be important for Paleogene cooling \cite{Kent2013a}. However, the higher 526--678~ppm values for the 15~Ma ``paleo-SEAI'' scenario illustrate that higher relief and a wet tropical climate significantly increase the efficiency of \COtwo consumption, especially when paired with high Ca+Mg lithologies. As such, arc-continent collisions in the tropics are likely more important for driving long-term changes in \pCOtwo than the eruption of flood basalts \cite{Macdonald2019a, Park2019a}.

@@

Marine \SrSr has overall been increasing since ca. 35~Ma \cite{McArthur2012a}. The traditional explanation for this trend is that it reflects increased weathering of radiogenic (i.e. high \SrSr) silicate rocks \cite{Raymo1988a}. Associated with this explanation is the proposal that increasing weathering of radiogenic silicate rocks in the Himalayas was the primary driver of Neogene cooling \cite{Raymo1992a, Edmond1992a}. It could be argued that increasing marine \SrSr is inconsistent with the hypothesis that increasing weathering of juvenile (i.e. low \SrSr) silicate rocks in the SEAI was an important driver of Neogene cooling. However, the globally averaged ratio of silicate weathering fluxes from radiogenic cratonic rocks versus juvenile arc lithologies can be at least partially decoupled from marine \SrSr via the regional weathering of isotopically unique lithologies. For example, unusually radiogenic carbonates are abundant in Himalayan strata, and it is estimated that $\sim$75\% of Sr coming from the Himalayas can be attributed to carbonate rather than silicate weathering \cite{Jacobson2002a, Quade2003a, Oliver2003a}. As such, there are challenges in interpreting the marine \SrSr record as a direct proxy for silicate weathering fluxes. Nevertheless, steadily increasing marine \SrSr is interrupted ca. 16~Ma when the slope of the \SrSr curve decreases \cite{McArthur2012a}. This decrease in slope has been attributed to coincident exhumation of relatively low \SrSr Outer Lesser Himalaya carbonates \cite{Myrow2015a, Colleps2018a}, but could also be at least partially driven by the emergence of low \SrSr lithologies in the SEAI during arc-continent collision.

@@

The long-term carbon-cycle mass balance can be perturbed via mechanisms that are disconnected from changes in volcanic degassing and silicate weathering rates. For example, sulphide oxidation coupled to carbonate dissolution could act as a source of \COtwo on million year time-scales \cite{Torres2014a}. Similarly, the weathering of sedimentary organic matter could serve as a source of \COtwo \cite{Hilton2014a}. On the other hand, enhanced burial of organic matter could be an important sink of \COtwo, as has been suggested in the Bengal Fan \cite{Galy2007a} and Taiwan \cite{Kao2014a}. An increased flux of dissolved nutrients derived from silicate weathering could even strengthen the biological carbon pump and sequester \COtwo as organic carbon in deep marine sediments \cite{Hartmann2013a}. The fluxes of \COtwo represented by these processes are not accounted for in our model framework, and could have been affected by emergence of the SEAI and/or Himalayan orogeny. \pCOtwo changes that result from these processes would be superimposed on \pCOtwo changes associated with evolving silicate weathering rates. However, our coupled weathering-climate model indicates that the \pCOtwo change associated with evolving silicate weathering rates driven by emergence of the SEAI is sufficient to explain the majority of Neogene cooling (Fig. \ref{fig:scenario_pCO2}). Without this emergence, \pCOtwo would have remained above the level necessary for the growth of Northern Hemisphere ice sheets.

### 2020-08-03 Manuscript/Supplementary_Information.tex
revision checkpoint

>>>>

\citet{Zhu2020a} reviewed three climate models (CESM2, CESM1, and CCSM4) more recent than the GFDL CM2.0 experiments, and found that they had equilibrium climate sensitivities (the long-term response of global mean surface temperature to a doubling of \pCOtwo) of 5.3, 4.2, and 3.2\degreesC respectively. Of these three climate models, they found that the equilibrium climate sensitivity of CESM1 (4.2\degreesC) was most in agreement with paleo-\pCOtwo and global mean surface temperature data. The GFDL CM2.0 experiments have a lower equilibrium climate sensitivity ($\sim$2\degreesC) than CESM1. This discrepancy indicates that our \pCOtwo estimates (Fig. \ref{fig:scenario_pCO2_all}) are biased toward higher values than what would have been obtained if we had used CESM1 as the climate model component. However, equilibrium climate sensitivity over the Neogene remains poorly constrained, with estimates ranging from 1.5--4.5\degreesC \citep{Zhu2020a}.
